---
title: "Binomial approximation to the Economist's prediction"
author: "Matthew Kay"
output: github_document
---

To build a Galton board for the Economist prediction, we need a binomial distribution that can approximate
its current prediction. In this document we'll find such a distribution.

Basically, what we want is a bin width, number of bins, and
a mean that we then throw into the Galton board renderer in [galton_board_quantile_ragg.Rmd](galton_board_quantile_ragg.Rmd).

## Setup

```{r setup_binom}
library(tidyverse)
library(Hmisc)
library(ggdist)

theme_set(theme_ggdist())
```

## Economist Data

Read in the data from the Economist, which can be obtained from the bottom of [this page](https://projects.economist.com/us-2020-forecast/president). It contains predicted probabilities of Trump and Biden achieving each number of electoral votes:

```{r eval=FALSE, include=FALSE}
download.file(
  "https://cdn.economistdatateam.com/us-2020-forecast/data/president/economist_model_output.zip", 
  "data/economist/model_output.zip"
)
unzip(
  "data/economist/model_output.zip", 
  files = "output/site_data//electoral_college_simulations.csv",
  exdir = "data/economist",
  junkpaths = TRUE,
)
unlink("data/economist/model_output.zip")
```

```{r}
df_econ = read.csv("data/economist/electoral_college_simulations.csv")
```

From this we can plot a histogram of Biden's predicted probability of achieving each number of electoral votes (with 270 being a majority):

```{r}
col_269 = "#1b9e77"

base_plot = df_econ %>%
  ggplot(aes(x = dem_ev)) +
  geom_histogram(aes(y = stat(density)), binwidth = 1, fill = "gray75") +
  geom_vline(xintercept = 269, color = col_269, size = 1) +
  annotate("label", x = 269, y = .05, label = "269", color = col_269, size = 3) +
  xlab("Electoral votes for Biden") +
  scale_y_continuous(breaks = NULL) +
  ylab("") +
  coord_cartesian(xlim = c(0, 540))

base_plot
```

Let's approximate this with a normal distribution by simply using the mean and variance of the predictive distribution:

```{r}
mean_ev = mean(df_econ$dem_ev)
var_ev = var(df_econ$dem_ev)
sd_ev = sd(df_econ$dem_ev)
```

And plot:

```{r}
col_normal = "#7570b3"
normal_density = tibble(x = 1:538, y = dnorm(x, mean_ev, sd_ev))

base_plot + 
  geom_line(aes(x = x, y = y), data = normal_density, color = col_normal, size = 1)
```

The Normal approximation looks decent. Let's continue down this train a bit and use the binomial approximation to the Normal by finding a binomial distribution with the same variance and then shifting its location to match the above Normal distribution. 2

We can use the fact that the variance of a Binomial distribution with probability 0.5 is equal to 1/4 the number of trials in the distribution to find the number of trials needed (which is the same as the height of the Galton board we would need to construct):

```{r}
bin_n = round(4 * var_ev)
bin_n
```

That would be a very large Galton board! Leaving that aside for a moment, let's see how well it approximates the distribution:

```{r}
col_binom = "#d95f02"
binom_mass = tibble(x = 0:538, y = dbinom(x + round(bin_n/2 - mean_ev), bin_n, 0.5))

base_plot + 
  geom_line(aes(x = x, y = y), data = normal_density, color = col_normal, size = 1) +
  geom_step(aes(x = x, y = y), data = binom_mass, color = col_binom, direction = "mid", size = 1)
```

The binomial distribution looks nearly identical to the Normal distribution here. However, like I said, it would require a very large Galton board to generate this distribution down to the single electoral vote level. So instead, let's use wider bins --- say, bins between 35 and 45 electoral votes wide. In fact, we'll pick a bin width that puts 269 as close to a bin boundary as we can, given the scaling/shifting of the distribution we will be doing: 

```{r}
candidate_bin_width = 35:45
candidate_bin_n = round(4 * var_ev / candidate_bin_width^2)
bin_width = candidate_bin_width[which.min(abs(
  (269/candidate_bin_width + candidate_bin_n/2 - mean_ev/candidate_bin_width) %% 1 - 0.5
))]
bin_width
```

That gives us a bin width of `r bin_width`, leading to...

```{r}
bin_n_small = round(4 * var_ev / bin_width^2)
bin_n_small
```

... a Galton board only `r bin_n_small` rows tall, which is much more manageable. Let's check on the approximation:

```{r}
binom_mass_small = tibble(
  x = 0:538, 
  y = dbinom(round(x/bin_width + bin_n_small/2 - mean_ev/bin_width), bin_n_small, 0.5) / bin_width
)

base_plot + 
  geom_line(aes(x = x, y = y), data = normal_density, color = col_normal, size = 1) +
  geom_step(aes(x = x, y = y), data = binom_mass_small, color = col_binom, direction = "mid", size = 1)

```

In case a bin boundary doesn't line up exactly at 269, we'll adjust the mean a little bit so that it does. Since
we've already picked a binning to minimize the distance between 269 and its closest bin boundary, this shouldn't
require us to fudge the mean too much:

```{r}
mean_ev_adj = mean_ev + ((269 + bin_n_small*bin_width/2 - mean_ev) %% bin_width - bin_width/2)
mean_ev_adj
```

That should make a bin boundary fall on 269:

```{r}
binom_mass_small = tibble(
  x = 0:538, 
  y = dbinom(round(x/bin_width + bin_n_small/2 - mean_ev_adj/bin_width), bin_n_small, 0.5) / bin_width
)

base_plot + 
  geom_line(aes(x = x, y = y), data = normal_density, color = col_normal, size = 1) +
  geom_step(aes(x = x, y = y), data = binom_mass_small, color = col_binom, direction = "mid", size = 1)
```
Thus, the final parameters for our approximation are:

```{r}
n_bin = bin_n_small
board_mean = mean_ev_adj
row_ratio = 2
model_name = "The Economist"

cat(paste0(
  "bins:      ", bin_n_small, "\n",
  "bin width: ", bin_width, "\n",
  "mean:      ", mean_ev_adj, "\n"
))
```

And we can use these quantiles for the dotplot:

```{r}
n_draw = 50
bin_values = as.vector(quantile(round((df_econ$dem_ev - mean_ev_adj)/bin_width + bin_n_small/2), ppoints(n_draw)))
bin_values
```

```{r child = 'galton_board_quantile_ragg.Rmd'}
```
